Narrative

### 2023-05-23 manuscript/manuscript.qmd
intro: improved long sentence in first paragraph

Estimating heterogeneity of effect sizes is usually considered the main purpose of meta-analysis in addition to estimating an average effect size. Heterogeneity of effect sizes (henceforth referred to as heterogeneity) refers to an effect size's sensitivity to variability in study design, or more specifically to sensitivity to differences in four study factors: 1) sample population, 2) settings, 3) treatment variables and 4) measurement variables (e.g., Campbell & Stanley, 2015). Heterogeneity is a primary outcome in meta-analysis because its presence signals the existence of moderators and so can be seen as an opportunity for theoretical development (Simons et al., 2017), because it affects the implementation of research by indicating that an intervention may not be equally effective under all conditions or for all people, and because its presence changes the interpretation of the average effect sized derived in meta-analysis from the true population effect size (under homogeneity) to the average of true subpopulation effect sizes (under heterogeneity). As such, estimating heterogeneity without bias is of central importance to meta-analysis (for a more extensive discussion on the importance of heterogeneity, see Olsson Collentine et al., 2020; Simons, 2017).

@@

Estimating heterogeneity in both the HV and bare-bones HS model essentially consists of comparing the observed variance in effect sizes with what would be expected from sampling error alone and then ascribing any excess observed variance to heterogeneity (Borenstein et al., 2010, Box 1; Hunter & Schmidt, 2015, p. XX). However, the process for doing so differs somewhat between methods. We apply the Restricted Maximum Likelihood (REML) estimator of heterogeneity when estimating heterogeneity in the Hedges and Vevea (1998) model, because it is the generally recommended heterogeneity estimator (Veroniki et al., 2016; Viechtbauer, 2005). Describing the REML estimator in detail is beyond the scope of this paper and we refer interested readers to the description in Veroniki et al., (2016). The Hunter and Schmidt (2015) model estimates heterogeneity as $\tau^2 = \sigma^2_r - \sigma^2_e$ where  $\sigma^2_r$ is the observed variance across correlations and $\sigma^2_e$ is the average sampling error variance across studies, computed as $ \frac{\Sigma[N_i \sigma^2_{ei}]}{\Sigma N_i}$ (p. 87, 2nd edition) and $\sigma^2_{ei}$ is the usual sampling variance for the Pearson correlation but with $\rho$ replaced with the sample size weighted average correlation $\bar{r} = \frac{\Sigma n_i r_i}{\Sigma n_i}$ instead of $r_i$ as is common. A generalized version of the HS estimator for effect sizes other than correlations can be found in Viechtbauer (2005; 2015).

>>>>

Estimating heterogeneity of effect sizes is usually considered the main purpose of meta-analysis in addition to estimating an average effect size. Heterogeneity of effect sizes (henceforth referred to as heterogeneity) refers to an effect size's sensitivity to variability in study design, or more specifically to sensitivity to differences in four study factors: 1) sample population, 2) settings, 3) treatment variables and 4) measurement variables (e.g., Campbell & Stanley, 2015). Heterogeneity is a primary outcome in meta-analysis because of several reasons. First, its presence signals the existence of moderators and so can be seen as an opportunity for theoretical development (Simons et al., 2017). Second, heterogeneity affects the implementation of research by indicating that an intervention may not be equally effective under all conditions or for all people. Third, the presence of heterogeneity changes the interpretation of the average effect sized derived in meta-analysis from the true population effect size (under homogeneity) to the average of true subpopulation effect sizes (under heterogeneity). As such, estimating heterogeneity with minimal bias is of central importance to meta-analysis (for a more extensive discussion on the importance of heterogeneity, see Olsson Collentine et al., 2020; Simons, 2017).

@@

Estimating heterogeneity in both the HV and bare-bones HS model essentially consists of comparing the observed variance in effect sizes with what would be expected from sampling error alone and then ascribing any excess observed variance to heterogeneity (Borenstein et al., 2010, Box 1; Hunter & Schmidt, 2015, p. XX). However, the process for doing so differs somewhat between methods. We apply the Restricted Maximum Likelihood (REML) estimator of heterogeneity when estimating heterogeneity in the Hedges and Vevea (1998) model, because it is the generally recommended heterogeneity estimator (Veroniki et al., 2016; Viechtbauer, 2005). Describing the REML estimator in detail is beyond the scope of this paper and we refer interested readers to the description in Veroniki et al. (2016). The Hunter and Schmidt (2015) model estimates heterogeneity as $\tau^2 = \sigma^2_r - \sigma^2_e$ where  $\sigma^2_r$ is the observed variance across correlations and $\sigma^2_e$ is the average sampling error variance across studies, computed as $ \frac{\Sigma[N_i \sigma^2_{ei}]}{\Sigma N_i}$ (p. 87, 2nd edition) and $\sigma^2_{ei}$ is the usual sampling variance for the Pearson correlation but with $\rho$ replaced with the sample size weighted average correlation $\bar{r} = \frac{\Sigma n_i r_i}{\Sigma n_i}$ instead of $r_i$ as is common. A generalized version of the HS estimator for effect sizes other than correlations can be found in Viechtbauer (2005; 2015).
